Explore expression of marker genes for tumor cell states

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
Apply suggestions from code review
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

1. The output from `aucell-singler-annotation.sh` is read in and annotations are lumped so that we can easily plot the top cell types. 
2. Clustering results from `evaluate-clusters.sh` is read in and subset to the desired parameters we would like to use for assigning clusters (leiden, modularity, resolution 0.5, nearest neighbors 20). 
3. The marker genes for tumor cell states and custom gene signatures are used to calculate the sum of all marker genes in each gene set for each library. 
4. UMAP embeddings for each library is pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.

@@

```{r}
cluster_classification_df <- umap_df |> 
  dplyr::mutate(
    cluster_classification = dplyr::case_when(
      library_id == "SCPCL000822" & cluster %in% c(1, 3, 4) ~ "tumor",
      library_id == "SCPCL000822" & cluster == 6 ~ "endothelial cell",
      # cluster 2 has predominantly chondrocytes and has MSC marker gene expression
      library_id == "SCPCL000822" & cluster == 2 ~ "chondrocyte",
      # clusters with high "immune" marker gene expression are macrophage
      library_id == "SCPCL000822" & cluster == 5 ~ "macrophage",
      # cluster 8 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000822" & cluster == 7 ~ "mesenchymal-like cell",
      library_id == "SCPCL000824" & cluster %in% c(2, 3, 4, 5, 6, 7) ~ "tumor", 
      library_id == "SCPCL000824" & cluster == 8 ~ "endothelial cell",
      # cluster 9 has high immune markers and is mostly macrohpages
      library_id == "SCPCL000824" & cluster == 9 ~ "macrophage",
      # cluster 1 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000824" & cluster == 1 ~ "mesenchymal-like cell",
      # if they didn't have high expression of any markers, we label them as unknown
      .default = "unknown"
    )
  )
```

@@

    # filter the sce to only those barcodes
    filtered_sce <- sce[ , tumor_cells]

    # grab pcs for clustering
    pcs <- reducedDim(filtered_sce)

@@

Now we will look at the marker gene expression across all clusters for each sample. 
First using a density plot and then a heatmap.

>>>>

1. The output from `aucell-singler-annotation.sh` is read in and annotations are lumped so that we can easily plot the top cell types. 
2. Clustering results from `evaluate-clusters.sh` is read in and subset to the desired parameters we would like to use for assigning clusters (leiden, modularity, resolution 0.5, nearest neighbors 20). 
3. The marker genes for tumor cell states and custom gene signatures are used to calculate the sum of all marker genes in each gene set for each library. 
4. UMAP embeddings for each library are pulled out into a data frame and combined with `SingleR` annotations, cluster assignments, and marker gene expression for all marker gene sets.

@@

```{r}
cluster_classification_df <- umap_df |> 
  dplyr::mutate(
    cluster_classification = dplyr::case_when(
      ############## Library SCPCL000822 ###############
      library_id == "SCPCL000822" & cluster %in% c(1, 3, 4) ~ "tumor",
      library_id == "SCPCL000822" & cluster == 6 ~ "endothelial cell",
      # cluster 2 has predominantly chondrocytes and has MSC marker gene expression
      library_id == "SCPCL000822" & cluster == 2 ~ "chondrocyte",
      # clusters with high "immune" marker gene expression are macrophage
      library_id == "SCPCL000822" & cluster == 5 ~ "macrophage",
      # cluster 8 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000822" & cluster == 7 ~ "mesenchymal-like cell",
      ############## Library SCPCL000824 ###############
      library_id == "SCPCL000824" & cluster %in% c(2, 3, 4, 5, 6, 7) ~ "tumor", 
      library_id == "SCPCL000824" & cluster == 8 ~ "endothelial cell",
      # cluster 9 has high immune markers and is mostly macrohpages
      library_id == "SCPCL000824" & cluster == 9 ~ "macrophage",
      # cluster 1 has high MSC marker gene expression and a mix of MSC like cell types
      library_id == "SCPCL000824" & cluster == 1 ~ "mesenchymal-like cell",
      # if they didn't have high expression of any markers, we label them as unknown
      .default = "unknown"
    )
  )
```

@@

    # filter the sce to only those barcodes
    tumor_sce <- sce[ , tumor_cells]

    # grab pcs for clustering
    pcs <- reducedDim(tumor_sce)

@@

Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then a heatmap.
